Eco-evolutionary dynamics and collective migration: implications for salmon metapopulation robustness

### 2017-09-19 Manuscript/FinalDraft_rev/ReviewResponses.tex

\opening{To the editorial committee,}

\setstretch{1.2}

\section*{Reviewer I}
\subsection*{Major comments:}
\noindent \textcolor{cyan}{
{\bf Comment 1.1} Equation (4), which describes the change in the population mean trait under dispersal and selection, is incorrect. In the second line of Eq (4), the immigrant term $(1-w_i) R_i [\mu_j]$ is independent of $\mu_i$, so that the derivative of this term is zero. This is as if selection did not change the allele frequencies/mean trait value of the immigrant subpopulation, which is obviously untrue (there is strong selection on the locally maladapted immigrants, favouring those who happen to be closer to the local optimum). One cannot apply Lande's equation without adapting it to the immigration model.
}

\noindent {\bf Response 1.1} We appreciate the Reviewer's concern, however note that equation 4 is written as
\begin{align}
  \label{eq:mu}
  \mu_i(t+1) &= w_i\mu_i(t) + (1-w_i)\mu_j(t) \\ \nonumber
  &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)] + (1-w_i)R_i[\mu_j(t)]  \right).
\end{align}
Because the third term on the right is the derivative of the log of the sum (as it must be for the discrete time version of the Lande equation; Lande \emph{Evolution}, 1976), the derivative of the second term within the logarithm is not zero, and both phenotypic means contribute to selection, such that
\begin{equation}
  \frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i] + (1-w_i)R_i[\mu_j]  \right) = \frac{\theta_i-\mu_i}{\left(\sigma ^2+\tau ^2\right) \left(1+\left(\frac{1}{w_i}-1\right) {\rm exp}\left[\frac{(\mu_i-\mu_j) (-2 \theta_i+\mu_i+\mu_j)}{2 \left(\sigma ^2+\tau ^2\right)}\right]\right)}.
\end{equation}

\noindent To clarify, we have made the following edits to the text:\\
\noindent {\bf Line 113:} \emph{The partial derivative in Eq. 4 determines how the mean trait changes through time due to natural selection [ref], which is proportional to the change in mean fitness with respect to $\mu_i$.
We note that the derivative is dependent on both $\mu_i$ and $\mu_j$ due to the influence of the logarithm on the sum.}

@@

\noindent \textcolor{cyan}{
{\bf Comment 1.2} The model with constant dispersal ($m$) is very much the same as the model of Ronce and Kirkpatrick (2001, Evolution), with the difference that the latter is fully deterministic and set in continuous time; but as long as the system attains an equilibrium rather than population cycles or chaos, continuous time should not make a difference. Ronce and Kirkpatrick focused on equilibria (such as the dominant vs subordinate state here ) and on local adaptation, not metapopulation resilience/stability, but did study the consequence of disturbances for switching between alternative equilibria (see the next point).
}

\noindent {\bf Response 1.2} We are thankful for the Reviewer in pointing out this reference, which we are embarrassed to admit to having overlooked. The model indeed has many similarities to our own, and we now acknowledge its importance and priority within the manuscript. The similarities in the resulting dynamics suggest that what we are observing in the two-population salmon system, as well as the implications that we derive for metapopulation robustness and recovery, may be  relevant and illustrative of a general phenomenon outside of the constraints introduced in the specific formulation of our model. Though as Reviewer I notes, our model while complementary to Ronce and Kirkpatrick 2001, focuses instead on the metapopulation effects (portfolio effects, resilience, and overall robustness) of the observed dynamics rather than a detailed analytical treatment of the dynamics themselves. We highlight areas of complementarity and extension in the revised manuscript.

\noindent We have made the following edits to the text:\\
\noindent {\bf Line 115:} \emph{This model formulation has parallels to that proposed by Ronce and Kirkpatrick [ref]), where habitat specialization evolves between two populations as a function of dispersal.
The largest difference between these approaches is that our framework treats trait evolution mechanistically (at some cost to analytical tractability).
Importantly, we show that the resulting dynamics are qualitatively similar, suggesting that the dynamical features present in both of these approaches have potentially widespread ramifications for the evolutionary dynamics of connected populations.}\\
\noindent {\bf Line 191} \emph{These dynamics are also observed in the Ronce and Kirkpatrick model, which they describe as a transition from symmetric to asymmetric states [ref].}

@@

\noindent \textcolor{cyan}{
{\bf Comment 1.3} The bifurcation the authors call a fold bifurcation is not a fold bifurcation. Figure 1 suggests a pitchfork bifurcation (which is degenerate but will appear in symmetric models). An alternative possibility is that there exists an asymmetric equilibrium with a "dominant state" and a "subordinate state" also at low m, next to the symmetric equilibrium shown at low m in figure 1, so that the system has two stable equilibria; here the "FB" point in figure 1 would be the point where the symmetric equilibrium is destabilized (this is the pattern found by Ronce and Kirkpatrick). This second alternative seems likely to me because the abrupt changes seen in figure 1 and also in figure 5 may well indicate an attractor switch. The nature of the bifurcation should be properly investigated (one can find Jacobians etc with numerical analysis for the deterministic system).
}

\noindent {\bf Response 1.3} Both Reviewers indicated that our labeling of the Fold Bifurcation was incorrect, and that the identity of the transition of the system from symmetric to asymmetric states was in fact a Pitchfork bifurcation. To our knowledge, the Pitchfork Bifurcation applies only to continuous time systems, and not to discrete maps. However, there is a qualitative (visual) analogue to the Pitchfork in discrete maps, and we think that we can be more precise in our description.

According to Kuznetsov (1998), there are three principle bifurcations in discrete maps, from which others are composed: the Fold, Flip (period-doubling), and Neimark-Sacker.
The Fold bifurcation is generally defined by the dominant eigenvalue $\lambda$ intersecting the unit circle in Real-Imaginary space at ${\rm Re}[\lambda]=+1$.
We now include a new figure S3 that shows that this is indeed the case for our system, and this was the basis on which we initially identified the bifurcation. Although we are not mathematicians, our understanding is that the fold bifurcation is a general attribute based on this eigenvalue condition, but as the Reviewers indicate, it is not a particularly clear or complete description of what is going on. 
A familiar, and more complicated bifurcation that occurs in maps with $\geq2$ parameters is the Discrete Cusp Bifurcation, which occurs when two fold bifurcations intersect at a cusp.
A one-dimensional transect through the cusp results in behavior similar to that observed for Pitchfork Bifurcations in vector fields (topological normal form for the discrete map: $x \mapsto x + \beta x - x^3$). At the cusp ($\beta=0$), the single fixed point crosses the intersecting fold bifurcations, and here it branches into 3 fixed points; two stable fixed points separated by an unstable fixed point.
We now identify the bifurcation as a Discrete Cusp bifurcation and mention that it displays dynamics that are visually similar to Pitchfork bifurcations in continuous time systems, which we hope will increase clarity and specificity.

@@

% Interestingly, because the system passes through a Pitchfork, or the Cusp at the cusp point, rather than one fold bifurcation at a time (which would result in similar stable-unstable-stable fixed points), the dependence of whether the system will veer to both states

As the Reviewers suggest, we investigated whether or not our system displays hysteresis, as does the system in the Ronce and Kirkpatrick model.
We now show that it does (figure S4), a feature characteristic of Cusp bifurcations in general.
This is more evidence that the dynamics described here and in Ronce and Kirkpatrick are perhaps general phenomena of spatially-linked eco-evolutionary systems.
This perspective is reinforced by the fact that altering parameters between the two sites (creating asymmetry) does not alter the qualitative nature of the system.

% we believe that our 4-dimensional discrete time system crosses a discrete cusp bifurcation at intermediate values of $m$. The discrete cusp bifurcation is characterized by the intersection of two branches of a fold bifurcation, of which the topological normal form is $x \mapsto f(x) = x + \beta_1 + \beta_2 x - x^3$. The eigenvalue evaluated at the steady state $\partial f(x)/\partial x|_{x=x^*}$ increases to unity on the unit circle at the cusp bifurcation $(\beta_1 = 0, \beta_2 = 0)$.

\noindent We have made the following edits to the text:\\
\noindent {\bf Line 115:} \emph{The decline in steady state densities is not gradual: as straying increases, the system crosses a discrete cusp bifurcation (DCB) [ref] whereby the single steady state for the metapopulation bifurcates into two basins of attraction: one at high biomass, and one at low biomass density (figure 1a, 2a).
Mean trait values for both populations bifurcate similarly (figure 1b). 
In discrete systems, the cusp bifurcation is defined by two fold bifurcations intersecting at a cusp point [ref], and is observed when the real part of the dominant eigenvalue of the Jacobian matrix crosses the unit circle at +1 (figure S3).
Visually, the dynamics are similar to those observed at a pitchfork bifurcation in continuous systems, where a single steady state gives rise to two steady states separated by an unstable fixed point.
}

@@

\noindent {\bf Response 1.4} We thank the Reviewer and apologize for our lack of clarity. We emphasize that our measure of recovery time is a numerical measurement, as eigenvalue-based (asymptotic) measures are not appropriate for the scale of disturbance that we introduce (extinction in some cases - far from the non-trivial fixed point(s) of the system). The procedure is now described in more detail within the text, and illustrated graphically in an additional supplemental figure (figure S1). In short, we track the recovering aggregate population $N_T=N_1+N_2$ until it is within a standard deviation around the \emph{updated} fixed point (thus accounting for fixed points that are altered by the disturbance). Note that this recovery time is specifically for the aggregate population, such that if a single population takes a long time to recovery relative to the other, the aggregate takes longer to recover as well. If $N_T$ remains within a standard deviation of the final fixed point for some period of time, it is counted as recovered. Accordingly, trajectories that cycle around the fixed point are not counted as recovered until the cycles decay. To ensure reproducibility, all code is provided online in the Github repository: https://github.com/jdyeakel/SalmonStrays.

\noindent We have made the following edits to the text:\\
\noindent {\bf Line 160:} \emph{Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation [ref], because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
Recovery time was calculated by initiating a disturbance at $t=t_d$, and monitoring $N_T(t_d+t)$ as $t\rightarrow T$, where $T$ is large. 
The aggregate was deemed recovered at $t_r$, such that recovery time was calculated as $t_r-t_d$, and recovery at $t=t_r$ was measured as the initial $t$ where $N_T(t) < {\rm SD}\left( N_T^* \right)$ for $t\in(t_r,T)$, where $\rm{SD}(\cdot)$ is standard deviation (illustrated in figure S2)}\\

@@

\noindent {\bf Response 1.5} We utilize the portfolio effect as a measurement to track the potential for the aggregate $N_T$ to buffer variance observed in the constituent populations $(N_1,N_2)$. Near the bifurcation, if $N_1$ and $N_2$ are more variable relative to $N_T$, then PE should be high, which is the principle goal of the measurement. 
However we agree that PE can change in response to changes to any of the four attributes from which it is composed (means of the aggregate, components; standard deviations of the aggregate, components), so it is sometimes hard to evaluate what is changing and what those changes mean relative to a single, descriptive measures of resilience such as recovery time. This is the primary motivation for incorporating both PE and recovery time in our analysis.

% While we agree with the Reviewer that the spike in PE at the bifurcation is not necessarily descriptive of a resilient population, the attributes that lead to an increase in the PE to the right of the bifurcation (higher $m$) also results in shorter recovery times, and is measuring the potential for the aggregate to buffer variance. We do not put much weight in the PE spike at the bifurcation as an indicator of metapopulation health

\subsection*{Minor comments:}
\noindent \textcolor{cyan}{
{\bf Minor Comment 1.1} ``straying" is not defined in the Introduction/main text. Since the model is not specific to salmon, the more general term "dispersal" could be used throughout.}

\noindent {\bf Minor Response 1.1} We now equate straying with dispersal early in the text (it is now also defined in the abstract).

\noindent We have made the following edits to the text:\\
\noindent {\bf Line 42:} \emph{The rate at which individuals stray, $m$, is in this case synonymous with dispersal and may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices [refs].}\\

@@

\noindent \textcolor{cyan}{
{\bf Minor Comment 1.2} The figure legends should specify the parameter values used, figure 2 should give a key translating colours to values, figure 3 should have numbers on the axes. Generally, the authors should pay attention to reproducibility.}

\noindent {\bf Minor Response 1.2} We agree and thank the Reviewer for pointing this out. We have made the following edits to the text:\\
\noindent {\bf Figure 1 caption} \emph{Unless otherwise indicated, the default parameter values used are: $r_{\rm max}=2$; $Z=0.5$; $\beta=0.001$; $\theta_1=5$; $\Delta\theta=5$; $\tau=1$; $\sigma=1$; $T=1\times10^5$.}\\

@@

\section*{Reviewer II}
\subsection*{Major comments}
\noindent \textcolor{cyan}{
{\bf Comment 2.1} My major criticism on this paper is that in the analysis presented, the bifurcation point is incorrectly identified and the claims based on this misidentification are most likely unsupported. What is presented as a Fold bifurcation is in reality a Pitchfork bifurcation. Pitchfork bifurcations are dependent on perfect symmetry in systems, something that is also true in the current case: The model as presented and analyzed is formulated in an entirely symmetrical way, which means that a Pitchfork bifurcation is de facto built into it.
}

\noindent {\bf Response 2.1} We thank the Reviewer for their comments and have now clarified the manuscript. Please see Responses 1.3 and 2.2, which address these concerns.

\noindent \textcolor{cyan}{
{\bf Comment 2.2} As it is, the occurrence of the bifurcation point as well as the ecological implications derived from it, are mostly a mathematical construction, brought about by the complete symmetry in the system. It is absolutely necessary to investigate the robustness of the results under scenarios where the symmetry is broken. The eco-evolutionary aspects of the model have currently no role in the observed dynamics, because they hinge on the symmetrical formulation.
I would be very curious to see this analysis extended under asymmetrical assumptions, for example in the productivity of the different locations or the intraspecific competition between populations.
\\
This manuscript can be improved by correcting the bifurcation analysis and by extending beyond perfect symmetry. The authors take a reasonable approach, using numerical analysis, which is fine, and I like the idea to focus on population robustness.  I understand that recruitment and population growth are not explicitly linked to environmental resources for reasons of mathematical simplicity. At the same time, I would prefer to see resources as an explicit variable in this model. This suggestion could directly be used to introduce asymmetries between two different locations and may be a feasible way to address the issues I raised above.
}

\noindent {\bf Response 2.2} We thank the Reviewer for their perspective and comments, and agree that we need to understand to what extent our results are generalizable.
% From our perspective, there are two ways in which our model incorporates symmetry: 1) the architectural symmetry of the populations between sites and 2) the symmetry in parameter values between sites. Regarding the first, both populations have the same growth and mortality functions, and both populations have the same functional forms that control straying. This architectural symmetry lies at the heart of our question, as we are exploring the dynamics of two similar populations that have similar constraints. We cannot alter this type of symmetry without vastly changing the questions that we are addressing.
In general, although discrete cusp bifurcations with pitchfork-like behaviors are a consequence of symmetry (defined by a system where $N_1\rightarrow-N_1$, $N_2\rightarrow-N_2$, $x_1\rightarrow-x_1$, $x_2\rightarrow-x_2$ without altering the equations), the general features of the bifurcation need not be altered as symmetry is broken.
Consider the topological normal form $\dot{x}=h + \beta x - x^3$, which is a continuous system with a perfect pitchfork bifurcation at $\beta = 0,~h=0$.
It is symmetric in the sense that $x\rightarrow-x$ without altering anything if and only if $h=0$.
If $h\neq0$, the system is asymmetric, however the pitchfork-like dynamic does not disappear but becomes an imperfect pitchfork where a saddle node bifurcation appears alongside a non-varying steady state.
This creates a parameter region where alternative steady states exist, separated by an unstable steady state.
As such, symmetry exists only in a perfect sense (and we describe below why we think our system already violates perfect symmetry), and the observed dynamics that we observe need not require perfect symmetry. 
However, we agree with the Reviewer that this begs investigation.

Our model treats many of the parameters of the model controlling vital rates to be the same between sites, primary examples being: $r_{\rm max}=2;~Z=0.5;~\beta=0.001$.
These parameters were held constant primarily to minimize the parameter space that we investigate (making an already long paper a bit shorter), but we acknowledge that this invites the question: are we investigating a mathematical knife-edge?
We suggest that our analysis is not merely a mathematical construction (interpreted here to mean dynamic behaviors that only occur within a thin slice of parameter space, unlikely to be realized in biological systems) for 3 reasons:

@@

1) We already inject process error into the dynamics. This process error (independent for both sites) serves to push the system away from the symmetrical vital rates $(r_{\rm max}, Z, \beta)$, and if there is a knife-edge of parameter space that results in the dynamics we observe, the process error should knock the system away from it. Despite these independent sources of error, their means are the same, and this warrants additional investigation as the Reviewer suggests.

2) We have included a modified analysis of the system to assess increasing asymmetry in the vital rates between sites. Asymmetry in parameter values is introduced as a parameter $\alpha$, where maximal growth at sites 1 and 2 are now  $r_{\rm max}(1)=r_{\rm max}(1+\tilde{rv}_1)$ and $r_{\rm max}(2)=r_{\rm max}(1+\tilde{rv}_2)$ where $rv_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $r_{\rm max}=2$. 
% Increasing $\alpha$ results in more widely diverging vital rates between sites.
Similarly the strength of density dependence is calculated at sites 1 and 2 as $\beta(1)=\beta(1+\tilde{rv}_1)$ and $\beta(2)=\beta(1+\tilde{rv}_2)$ where $\tilde{rv}_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $\beta=0.001$.
Thus as asymmetry ($\alpha$) increases, so do the differences in vital rates between sites.

### 2017-09-19 Manuscript/FinalDraft_rev/msv6.tex
final review response

% \bibliographystyle{prsb}

@@

\noindent where the first two components determine the mixed normal average of the aggregated local and remote populations.
The partial derivative in the Eq. \ref{eq:mu} determines how the mean trait changes through time due to natural selection \citep{Lande:1976ga}, which is proportional to the change in mean fitness with respect to $\mu_i$.
We note that the derivative is dependent on both $\mu_i$ and $\mu_j$ due to the influence of the logarithm on the sum.
%Comparison to the Ronce and Kirkpatrick paper
This model formulation has parallels to that proposed by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where habitat specialization evolves between two populations as a function of dispersal.
The largest difference between these approaches is that our framework treats trait evolution mechanistically at some cost to analytical tractability.
Importantly, we show that the resulting dynamics are qualitatively similar, suggesting that the dynamical features present in both of these approaches have potentially widespread ramifications for the evolutionary dynamics of spatially connected populations, regardless of the specific formulation proposed.
\\

@@

Numerically estimating the time that it takes for a perturbed system to recover also permits a more detailed perspective of metapopulation fragility.
For example, if populations settle to alternative stable states, comparing recovery times after a disturbance applied to individual populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery. % (alternative stable states in our model requiring one population to be high-density and one low-density)
We measured recovery time following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if populations have the same density); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness and by extension persistence.
\\

@@

\noindent{\bf (a) Nonlinear effects of straying on metapopulation robustness} \\
\noindent Regardless of density dependence, straying lowers steady state densities for both populations by (\emph{i}) the donor population losing locally-adapted individuals to the recipient population and (\emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population (Fig. \ref{fig:traj}).
This prediction is in accordance with observations from natural populations \citep{Bett:2017ha}. %, regardless of trait heritability and variance or habitat heterogeneity
The decline in steady state densities is not gradual: as straying increases, the system crosses a pitchfork bifurcation (PFB) \citep{AleksandrovichKuznetsov:1995p2580} whereby the single steady state for the metapopulation bifurcates into two basins of attraction: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
In discrete systems, a pitchfork bifurcation is a specific case of the cusp bifurcation (where two fold bifurcations intersect at a cusp) \citep{AleksandrovichKuznetsov:1995p2580}, and occurs when the real part of the dominant eigenvalue of the Jacobian matrix crosses the unit circle at +1 (figure \ref{fig:eigs}).
Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b).

Above the threshold straying rate defined by the PFB, there are two alternative eco-evolutionary states: the \emph{dominant state} population will have a higher density and greater degree of local adaptation (smaller trait offset from the local optimum), while the \emph{subordinate state} population will have lower density with maladapted trait values (larger trait offset from the local optimum). 
Hysteresis is observed to occur at this transition, such that the single steady state regime cannot easily be recovered by reducing straying after the system attains alternative steady states (figure \ref{fig:hysteresis}).
These dynamics are also observed in the Ronce and Kirkpatrick model, where populations are described as transitioning between symmetric to asymmetric states \citep{Ronce:2001dp}.
Whether a specific population goes to one state or the other in our model is random, due to a small amount of introduced variance in the initial conditions.

Trait heritability $h^2$ has a large effect on the degree to which straying affects both the aggregate population steady state density ($N^*_T$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N^*=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
Greater trait heritability results in a steeper decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N^*$.
Lower trait heritability has little effect on the total biomass density but contrastingly large effects on $\Delta N^*$.
The pitchfork bifurcation (the black line in Figs. \ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with lower heritability (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
Although trait heritability among salmonids is variable, most life history traits have an $h^2 <0.5$ \citep{Carlson:2008hl}, and we largely focus our efforts on that range.

%Critical slowing down
As the pitchfork bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations.
This amplification in variance is the product of a dynamical process known as \emph{critical slowing down} that can occur near bifurcations \citep{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \citep{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br,Krkosek:2014ch}.
For larger values of $m$ (to the right of the pitchfork bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T^*$ increases.
The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high (figure \ref{fig:PE}c).

@@

Straying had non-linear impacts on the recovery time of populations. 
When the dominant state (well adapted and high density) is wiped out, high levels of straying allow it to recover quickly (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the recovering population (figure \ref{fig:relaxtraj_hdlh}).
Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is enough time for the trait distribution to move back towards the trait optimum of the subordinate state population.
In contrast, when the subordinate state population (maladapted and low density) is wiped out, recovery rates are fastest at low to intermediate levels of straying.
Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying, while near the onset of the pitchfork bifurcation, recovery time increases explosively and this is -- as the name implies -- characteristic of \emph{slowing} dynamics that occur near critical transitions \citep{Scheffer:2009gg,Kuehn:2010p2591}.\\

@@

%Recovery times
Density-dependent straying alters these recovery times (figure \ref{fig:relax}b). 
First, in comparison with constant stray rates, density-dependent straying made recovery more rapid at elevated stray rates when both populations collapsed and when the subordinate population was extirpated.
%Recovery times
%The recovery times for systems with constant and density-dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
At low straying rates, near-collapse of both populations resulted in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
As trait heritability increased, the metapopulation always recovered more quickly if the small population was lost (figure \ref{fig:relax_highh}).
% There was greater similarity in the recovery times for constant and density-dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
The lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:thetaPE}).
% This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
In tandem, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, it does appear to promote robustness, particularly when the aggregate biomass is low and straying is correspondingly high.

>>>>

\bibliographystyle{prsb}

@@

\noindent where the first two components determine the mixed normal average of the aggregated local and remote populations.
The partial derivative in Eq. \ref{eq:mu} determines how the mean trait changes through time due to natural selection \citep{Lande:1976ga}, which is proportional to the change in mean fitness with respect to $\mu_i$.
We note that the derivative is dependent on both $\mu_i$ and $\mu_j$ due to the influence of the logarithm on the sum.
%Comparison to the Ronce and Kirkpatrick paper
This model formulation has parallels to that proposed by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where habitat specialization evolves between two populations as a function of dispersal.
The largest difference between these approaches is that our framework treats trait evolution mechanistically at some cost to analytical tractability.
Importantly, we show that the resulting dynamics are qualitatively similar, suggesting that the dynamical features present in both of these approaches have potentially widespread ramifications for the evolutionary dynamics of spatially connected populations, regardless of the specific formulation proposed.
\\

@@

Numerically estimating the time that it takes for a perturbed system to recover also permits a more detailed perspective of metapopulation fragility.
For example, if populations settle to alternative stable states (one at low- and one at high-density), comparing recovery times after a disturbance applied to individual populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery. % (alternative stable states in our model requiring one population to be high-density and one low-density)
We measured recovery time following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if populations have the same density); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness.
\\

@@

\noindent{\bf (a) Nonlinear effects of straying on metapopulation robustness} \\
\noindent Regardless of density dependence, straying lowers steady state densities for both populations by (\emph{i}) the donor population losing locally-adapted individuals to the recipient population and (\emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population (Fig. \ref{fig:traj}).
This prediction is in accordance with observations from natural populations \citep{Bett:2017ha}. %, regardless of trait heritability and variance or habitat heterogeneity
The decline in steady state densities is not gradual: as straying increases, the system crosses a discrete cusp bifurcation (DCB) \citep{AleksandrovichKuznetsov:1995p2580} whereby the single steady state for the metapopulation bifurcates into two basins of attraction: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b). 
In discrete systems, the cusp bifurcation is defined by two fold bifurcations intersecting at a cusp point \citep{AleksandrovichKuznetsov:1995p2580}, and is observed when the real part of the dominant eigenvalue of the Jacobian matrix crosses the unit circle at +1 (figure \ref{fig:eigs}).
Visually, the dynamics are similar to those observed at a pitchfork bifurcation in continuous systems, where a single steady state gives rise to two steady states separated by an unstable fixed point.

Above the threshold straying rate defined by the DCB, there are two alternative eco-evolutionary states: the \emph{dominant state} population will have a higher density and greater degree of local adaptation (smaller trait offset from the local optimum), while the \emph{subordinate state} population will have lower density with maladapted trait values (larger trait offset from the local optimum). 
Hysteresis is observed to occur at this transition, such that the single steady state regime cannot easily be recovered by reducing straying after the system attains alternative steady states (figure \ref{fig:hysteresis}).
These dynamics are also observed in the Ronce and Kirkpatrick model, where populations are described as transitioning between symmetric to asymmetric states \citep{Ronce:2001dp}.
Whether a specific population goes to one state or the other in our model is random, due to a small amount of introduced variance in the initial conditions.

Trait heritability $h^2$ has a large effect on the degree to which straying affects both the aggregate population steady state density ($N^*_T$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N^*=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
Greater trait heritability results in a steeper decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N^*$.
Lower trait heritability has little effect on the total biomass density but contrastingly large effects on $\Delta N^*$.
The cusp bifurcation (the black line in Figs. \ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with lower heritability (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
Although trait heritability among salmonids is variable, most life history traits have an $h^2 <0.5$ \citep{Carlson:2008hl}, and we largely focus our efforts on that range.

%Critical slowing down
As the cusp bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations.
This amplification in variance is the product of a dynamical process known as \emph{critical slowing down} that can occur near bifurcations \citep{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \citep{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br,Krkosek:2014ch}.
For larger values of $m$ (to the right of the cusp bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T^*$ increases.
The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high (figure \ref{fig:PE}c).

@@

Straying had non-linear impacts on the recovery time of populations. 
When the dominant state (well adapted and high density) is wiped out, high rates of straying allow it to recover quickly (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the recovering population (figure \ref{fig:relaxtraj_hdlh}).
Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is time for the trait distribution to move back towards the trait optimum of the subordinate state population.
In contrast, when the subordinate state population (maladapted and low density) is wiped out, recovery rates are fastest at low to intermediate levels of straying.
Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying due to reduced mixing of maladaptive phenotypes.
Near the onset of the cusp bifurcation, recovery time increases explosively, however this is -- as the name implies -- characteristic of \emph{slow} dynamics occurring near critical transitions \citep{Scheffer:2009gg,Kuehn:2010p2591}.

%Lowering Phenotypic diversity with increasing straying rate
Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i^*-\mu_j^*|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
The loss of phenotypic diversity with increased straying is greater if trait heritability is low because traits take longer to go back to their local optima than they do when heritability is high. 
Hence straying counters the effect of diversifying local adaptation. 
% The loss of phenotypic diversity with increased straying is greater if trait heritability is low because the selective forces acting against trait means far from the local optima are lessened. % with decreased $h^2$.
Less intuitively, we observe a discrete jump towards low phenotypic diversity as the cusp bifurcation is crossed (figure \ref{fig:traitdiff}).
Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, entering this dynamic regime also results in a substantial decline in phenotypic diversity, which may have less predictable adverse effects on the population.\\

@@

%Recovery times
Density-dependent straying alters these recovery times (figure \ref{fig:relax}b; note difference in x-axis scale). 
First, in comparison with constant stray rates, density-dependent straying lowered recovery times at elevated straying rates for near-collapse of both populations and extirpation of the subordinate population.
%Recovery times
%The recovery times for systems with constant and density-dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
In contrast, at low straying rates, near-collapse of both populations required more time to recover. %whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
Trait heritability had a large effect on recovery times, with near-collapse requiring a more protracted time to recover in the alternative steady state regime (figure \ref{fig:relax_highh}).
% As trait heritability increased, the metapopulation always recovered more quickly if the subordinate population was lost (figure \ref{fig:relax_highh}).
% There was greater similarity in the recovery times for constant and density-dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
In general, the lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:thetaPE}a).
% This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
Together, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, it does appear to promote robustness, particularly in the case of near-collapse of both populations when straying is high.
\\
